Insecticide resistance evolution with mixtures and sequences : a model-based graphical introduction and summary

### 2017-08-08 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
comments from Hilary
paper2 writing

Modelling studies have investigated the evolution of insecticide resistance in insecticide mixtures including in a public health context e.g. [@Curtis1985][@Mani1985][@Roush1989] but much of the work was done more than 20 years ago and there remained some confusion about the results [@Levick2017]. In an earlier paper [@Levick2017] we described the technical details of a flexible model used to investigate the relative benefits of mixtures and sequences. Here we provide an accessible summary of the model and use selected parameter values to describe mechanistically how the evolution of resistance is influenced by different inputs. This mechanistic understanding can contribute to the debate on the relative merits of different insecticide strategies extending existing frameworks [@Roush1989][@IRAC2011][@FAO2012][@WHO2012].

Our modelling approach focuses on the change in frequency of a single resistance gene for each insecticide. It is likely that genes conferring resistance are present in populations even prior to exposure to novel insecticides thus leading to the potential for selection [@Liu2015].

@@

Relatedly, when there are costs of resistance, the frequency of resistance for insecticides that are not being used is likely to decline. Thus the frequency of resistance for the first insecticide in a sequence, after it has been replaced, may decline to a level where it becomes operationally useful again. As such a repeated sequential strategy could offer benefits over mixtures not considered here. To incorporat this one would need to use a more detailed metric than time-to-resistance.

@@

The development of insecticide resistance in operational field settings is subject to considerable uncertainty [@Ranson2016]. The model we present here is able to cover the range of that uncertainty, yet in this paper, to develop an understanding of the key mechanisms we restrict our explorations to a limited region of parameter space (e.g. holding most inputs constant at intermediate values while modifying single inputs in isolation). Thus there is the potential that the system may behave differently in the field than we describe here. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. This previous analysis also highlighted the relative importance of insecticide effectiveness and exposure in determining the relative performance of mixtures and sequences. In addition we provide an on-line user interface to the model [@South2017a] and enabling readers to investigate model behaviour in areas of parameter space we have not shown.

@@

>>>>

Modelling studies have investigated the evolution of insecticide resistance in insecticide mixtures including in a public health context e.g. [@Curtis1985][@Mani1985][@Roush1989] but much of the work was done more than 20 years ago and there remained some confusion about the results [@Levick2017]. In an earlier paper [@Levick2017] we described the technical details of a flexible model used to investigate the relative benefits of mixtures and sequences. Here we provide an accessible summary of the model and use selected parameter values to describe mechanistically how the evolution of resistance can be expected to change in different scenarios. This mechanistic understanding can contribute to the debate on the relative merits of different insecticide strategies extending existing frameworks [@Roush1989][@IRAC2011][@FAO2012][@WHO2012].

Our modelling approach focuses on the change in frequency of a single resistance gene for each insecticide. We assume both that genes conferring resistance are already present in the population and that resistance to each insecticide is coded by a single gene. These assumptions are consistent with previous modelling work. In support of the first assumption insecticide resistance has been termed 'pre-adaptive' meaning that resistance alleles are present in very low numbers even prior to exposure to novel insecticides [@Ffrench-Constant2013][@Liu2015] as has been documented in blowflies [Hartley2006]. In support of using a single gene representation of resistance evidence suggests that whilst polygenic resistance is common, control failures, particularly those due to target site resistance, are mostly due to single major genes [@Ffrench-Constant2013].

@@

Relatedly, when there are costs of resistance, the frequency of resistance for insecticides that are not being used is likely to decline. Thus the frequency of resistance for the first insecticide in a sequence, after it has been replaced, may decline to a level where it becomes operationally useful again. As such a repeated sequential strategy could offer benefits over mixtures not considered here. To incorporate this one would need to use a more detailed metric than time-to-resistance.

@@

There is much evidence of the involvement of multiple genes in resistance mechanisms. Our model, in common with most previous ones, just follows a single gene per insecticide. In most cases we believe this is reasonable. Target-site resistance has been shown to be mostly determined by single genes, and control failures in general have mostly been attributed to the effect of single major genes[]. Control doses of insecticides in the field are likely to lead to the selection of single major genes whereas the lower doses used to select laboratory strains are more likely to lead to polygenic resistance [@Ffrench-Constant2013]. To represent polygenic resistance a different 'quantitative genetic' approach would be required. A quantitive genetic model would likely lead to a more rapid initial evolution of resistance, a slower middle phase and a continuing increase. If polygenic resistance is shown in future to be more of a problem then these results could be compared to those of quantitative genetic models.

Even if the assumption that single genes are responsible for control failures is correct there is still considerable other uncertainty around the development of insecticide resistance in operational field settings [@Ranson2016]. The model we present here is able to cover much of the range of that uncertainty, yet in this paper, to develop an understanding of the key mechanisms we restrict our explorations to a limited region of parameter space (e.g. holding most inputs constant at intermediate values while modifying single inputs in isolation). Thus there is the potential that the system may behave differently in the field than we describe here. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. This previous analysis also highlighted the relative importance of insecticide effectiveness and exposure in determining the relative performance of mixtures and sequences. In addition we provide an on-line user interface to the model [@South2017a] and enabling readers to investigate model behaviour in areas of parameter space we have not shown.

@@

_Can you give some indications of the type of data that should be collected to help paramertise the model?_
